\pkg{DNest4}: Diffusive Nested Sampling in \proglang{C++} and \proglang{Python}

### 2016-05-10 paper/ms.tex
DOn't need escape character in code block
List of member functions required

%% an abstract and keywords
\Abstract{In probabilistic (Bayesian) inferences, we typically want to compute
properties of the posterior distribution, describing knowledge of
unknown quantities in the context of a particular dataset and the assumed
prior information. The marginal likelihood (also known as the ``evidence'')
is a key quantity in Bayesian model averaging. The Diffusive Nested Sampling algorithm, a variant of Nested Sampling, is a powerful tool for generating
posterior samples and estimating marginal likelihoods. It is effective at
solving complex problems including many where the posterior distribution is
multimodal or has strong dependencies between variables. The key requirement
is that the likelihood function be fast to evaluate.
\pkg{DNest4} is a multi-threaded implementation of this algorithm in
modern \proglang{C++} (specifically, \proglang{C++11}),
along with associated utilities including: i)
a \pkg{Python} package allowing basic use without \proglang{C++} coding;
ii) \pkg{RJObject}, a class template for finite mixture models;
iii) Experimental support for models implemented in \proglang{Julia};
iv) An experimental tool \proglang{Python} tool to
generate \pkg{C++} code from a model specification.
In this paper we demonstrate \pkg{DNest4} usage through examples including
simple Bayesian data analysis, finite mixture models, and Approximate
Bayesian Computation (ABC).
}

@@

In Nested Sampling, the target distribution is not the posterior but rather
the constrained prior. Hence, the appropriate acceptance probability is
\begin{align}
\alpha &= \min\left[1,
\frac{q(\params'|\params)}{q(\params | \params')}
\frac{\pi(\params')}{\pi(\params)}
\mathds{1}\left(L(\params') > L^*\right)
\right]
\end{align}
where $L^*$ is the threshold value of the likelihood.

@@

\section{Implementing models}
The ``classic'' method of implementing models in \pkg{DNest4} is by
writing a \proglang{C++} class, an object of which represents a
point in your parameter space.

To demonstrate, consider a simple linear regression situation where the
sampling distribution is
\begin{align}
y_i | m, b, \sigma &\sim \textnormal{Normal}(mx_i + b, \sigma^2)
\end{align}
and the priors are
\begin{align}
m &\sim \textnormal{Normal}(0, 1000^2)\\
b &\sim \textnormal{Normal}(0, 1000^2)\\
\ln\sigma &\sim \textnormal{Uniform}(-10, 10)
\end{align}
These are naive diffuse priors and we do not claim any special status for them.

@@

\begin{CodeChunk}
\begin{CodeInput}
void StraightLine::from_prior(DNest4::RNG\& rng)
{
   // Naive diffuse prior
   m = 1E3*rng.randn();
   b = 1E3*rng.randn();

>>>>

%% an abstract and keywords
\Abstract{In probabilistic (Bayesian) inferences, we typically want to compute
properties of the posterior distribution, describing knowledge of
unknown quantities in the context of a particular dataset and the assumed
prior information. The marginal likelihood (also known as the ``evidence'')
is a key quantity in Bayesian model averaging. The Diffusive Nested Sampling algorithm, a variant of Nested Sampling, is a powerful tool for generating
posterior samples and estimating marginal likelihoods. It is effective at
solving complex problems including many where the posterior distribution is
multimodal or has strong dependencies between variables. The key requirement
is that the likelihood function be fast to evaluate.
\pkg{DNest4} is a multi-threaded implementation of this algorithm in
modern \proglang{C++} (specifically, \proglang{C++11}),
along with associated utilities including: i)
a \pkg{Python} package allowing basic use without \proglang{C++} coding;
ii) \pkg{RJObject}, a class template for finite mixture models;
iii) Experimental support for models implemented in \proglang{Julia};
iv) An experimental \proglang{Python} tool to
generate \pkg{C++} code from a model specification.
In this paper we demonstrate \pkg{DNest4} usage through examples including
simple Bayesian data analysis, finite mixture models, and Approximate
Bayesian Computation (ABC).
}

@@

In Nested Sampling, the target distribution is not the posterior but rather
the constrained prior. Hence, the appropriate acceptance probability is
\begin{align}
\alpha &= \min\left[1,
\frac{q(\params'|\params)}{q(\params | \params')}
\frac{\pi(\params')}{\pi(\params)}
\mathds{1}\left(L(\params') > L^*\right)
\right]\\
&= \min\left[1,
e^{\ln(H)}
\mathds{1}\left(L(\params') > L^*\right)
\right]\label{eqn:log_hastings}
\end{align}
where $L^*$ is the threshold value of the likelihood.

@@

\section{Implementing models}
The ``classic'' method of implementing models in \pkg{DNest4} is by
writing a \proglang{C++} class, an object of which represents a
point in your parameter space.
To demonstrate, consider a simple linear regression situation where the
sampling distribution is
\begin{align}
y_i | m, b, \sigma &\sim \textnormal{Normal}(mx_i + b, \sigma^2)
\end{align}
and the priors are
\begin{align}
m &\sim \textnormal{Normal}(0, 1000^2)\\
b &\sim \textnormal{Normal}(0, 1000^2)\\
\ln\sigma &\sim \textnormal{Uniform}(-10, 10)
\end{align}
These are naive diffuse priors and we do not claim any special status for them.

Running \pkg{DNest4} on this problem requires a class, which we call
{\tt StraightLine}, which has the following member functions:
\begin{itemize}
\item \code{void from\_prior(DNest4::RNG\& rng)}, which generates parameter
        values from the prior;
\item \code{double perturb(DNest4::RNG\& rng)}, which proposes a
        change to the parameter values, and returns $\ln(H)$ as defined in
        Equation~\ref{eqn:log_hastings};
\item \code{double log\_likelihood() const}, which evaluates the log of
        the likelihood function;
\item \code{void print(std::ostream\& out) const}, which prints parameters of
        interest to the given output stream; and
\item \code{std::string description() const}, which returns a \proglang{C++}
        string naming the parameters printed by \code{print(std::ostream\&) const}.
\end{itemize}

@@

\begin{CodeChunk}
\begin{CodeInput}
void StraightLine::from_prior(DNest4::RNG& rng)
{
   // Naive diffuse prior
   m = 1E3*rng.randn();
   b = 1E3*rng.randn();
